---
title: "Hierarchical Model"
author: "Alice Jiang"
date: "11/17/2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
#library(nlme)
library(lme4)
library(ggplot2)
library(dplyr)
library(sjPlot) #for plotting lmer and glmer mods
library(groupdata2)
library(hydroGOF) # rmse()
```

```{r}
mydata <- read_csv("covid.csv") 
mydata <- mydata %>%
  select(-c(1:2,29:31))
mydata <- mydata[is.finite(mydata$covid_death_rate_log),]
mydata$name <- as.factor(mydata$name)
mydata$region <- as.factor(mydata$region)
mydata$state <- as.factor(mydata$state)
mydata$govParty <- as.factor(mydata$govParty)
mydata$metro_area <- as.factor(mydata$metro_area)
mydata$main_econ <- as.factor(mydata$main_econ)
mydata$GQ_ESTIMATES_2019 <- log(mydata$GQ_ESTIMATES_2019) 
mydata <- mydata %>%
  filter(GQ_ESTIMATES_2019 != '-Inf') %>%
  mutate(density = total_population/area_sqmi)
head(mydata)
#dim(mydata)
#death_rate <- exp(mydata$covid_death_rate_log)*100
#summary(death_rate)
```

```{#r}
summary(mydata)
summary(mydata$name)
```
$name$: county
demographic factors: $total\_population$, $percent\_fair\_or\_poor\_health$, $per\_capita\_income$
social factors: $percent\_uninsured$, $percent\_minorities$, $percent\_below\_poverty$, $percentile\_rank\_social\_vulnerability$
geographical factors: $region$ and $state$
economic factors: $main\_econ$ and $metro\_area$
political factors: $pro\_Trump$ and $govParty$

$GQ\_ESTIMATES\_2019$: the total number of county residents living in group quarters in 2019 (e.g. nursing homes, residential treatment centers, student housing, religious convents/monasteries, correctional facilities, and hospitals)

```{r}
ggplot(mydata, aes(x=covid_death_rate_log)) +
  geom_histogram(fill = "cornflowerblue", color = "white") +
  facet_wrap(~region, ncol = 4) +
  labs(title = "Death Rates by Region")

ggplot(mydata, aes(x=covid_death_rate_log)) +
  geom_histogram(fill = "cornflowerblue", color = "white") +
  facet_wrap(~metro_area, ncol = 3) +
  labs(title = "Death Rates by Metro Area")

ggplot(mydata, aes(x=covid_death_rate_log)) +
  geom_histogram(fill = "cornflowerblue", color = "white") +
  facet_wrap(~govParty, ncol = 2) +
  labs(title = "Death Rates by govParty")
```

# EDA
```{r}
ggplot(mydata, aes(x = region, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by Region")

ggplot(mydata, aes(x = metro_area, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by Metro Area")

ggplot(mydata, aes(x = govParty, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by govParty")

ggplot(mydata, aes(x = main_econ, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by Main Economy")
```
Northeast seems to have higher log(death rates).
metro_area, govParty, main_econ do not seem to have significant impacts on death rates.

```{r}
ggplot(mydata, aes(x = GQ_ESTIMATES_2019, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = per_capita_income, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = proTrump, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_below_poverty, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_fair_or_poor_health, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_minorities, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_uninsured, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percentile_rank_social_vulnerability, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm") 
```

*Positive association between group quarters and log(death rates)
*Negative association between proTrump and log(death rates).
Positive association between percent_below_poverty and log(death rates)
*Positive association between percent_fair_or_poor_health and log(death rates)
*Positive association between percent_minorities and log(death rates)
Positive association between percentile_rank_social_vulnerability and log(death rates)
No obvious relationship between log(death rates) and percent_uninsured and per capita income

```{#r}
# subset counties -> Columbia in various states
plotdata <- mydata %>%
    filter(name == "Columbia")
# basic Cleveland plot 
ggplot(plotdata, aes(x=name, y=covid_death_rate_log)) +
    geom_point()

# subset counties -> Franklin in various states
plotdata <- mydata %>%
    filter(name == "Franklin")
# basic Cleveland plot 
ggplot(plotdata, aes(x=name, y=covid_death_rate_log)) +
    geom_point()
```

```{#r}
# plot death rates by group quarters, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=GQ_ESTIMATES_2019, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=GQ_ESTIMATES_2019, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=GQ_ESTIMATES_2019, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=GQ_ESTIMATES_2019, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{#r}
# plot death rates by percent below poverty, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=percent_below_poverty, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_below_poverty, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_below_poverty, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_below_poverty, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{#r}
# plot death rates by proTrump, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=proTrump, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=proTrump, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=proTrump, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=proTrump, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{#r}
# plot death rates by percent_fair_or_poor_health, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=percent_fair_or_poor_health, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_fair_or_poor_health, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_fair_or_poor_health, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_fair_or_poor_health, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{#r}
# plot death rates by percent_minorities, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=percent_minorities, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_minorities, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_minorities, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_minorities, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{#r}
# plot death rates by percentile_rank_social_vulnerability, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=percentile_rank_social_vulnerability, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percentile_rank_social_vulnerability, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percentile_rank_social_vulnerability, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percentile_rank_social_vulnerability, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{#r}
# plot death rates by percent_uninsured, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=percent_uninsured, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_uninsured, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_uninsured, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=percent_uninsured, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{#r}
# plot death rates by per capita income, for each region, metro_area, main_econ, govParty
ggplot(mydata, aes(x=per_capita_income, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=per_capita_income, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~metro_area) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=per_capita_income, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~main_econ) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")

ggplot(mydata, aes(x=per_capita_income, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="blue") +
   facet_wrap(~govParty) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

# Linear Mixed Model 
```{r}
dotchart(sort(xtabs(~ mydata$name)), cex=0.7) 
dotchart(sort(xtabs(~ mydata$state)), cex=0.7) 
dotchart(sort(xtabs(~ mydata$region)), cex=0.7)
```
Distribution of potential random effects variables. There are many $name$ (county), each only contribute one instances). There's many states, with a pretty skewed distribution -> the state effects can be worrying. Finally, there's only few 4 regions.

```{r}
model1 <- lmer(covid_death_rate_log ~ (1|name) + (1|state) + (1|region), data=mydata)
summary(model1)
```
The estimated variances for the random effects. For $name$, the optimal variance is  0.00123, which is very close to zero: the model gains nothing by attributing the response to county variation. So $name$ isn’t a significant effect in the model. Hence we can drop the random effect for $name$. But for $state$ and $region$, the choice of state and region introduce a larger variance. 

```{r}
model2 <- lmer(covid_death_rate_log ~ (1|state) + (1|region), data=mydata)
summary(model2)
```
Dropping random effect for $name$ makes almost no difference to the model fit.

```{#r}
model3 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_fair_or_poor_health + percent_minorities + percent_uninsured + percentile_rank_social_vulnerability + govParty + main_econ + metro_area + (1|state) + (1|region), data=mydata)

model4 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_fair_or_poor_health + percent_minorities + percent_uninsured + percentile_rank_social_vulnerability + govParty + main_econ + (1|state) + (1|region), data=mydata)

model5 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_fair_or_poor_health + percent_minorities + percent_uninsured + percentile_rank_social_vulnerability + govParty + metro_area + (1|state) + (1|region), data=mydata)

anova(model3, model4)
anova(model3, model5)
```
metro_area and main_econ significant

```{#r}
model6 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_fair_or_poor_health + percent_minorities + percent_uninsured + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

anova(model3, model6)
```
govParty not significant

```{#r}
model7 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_fair_or_poor_health + percent_minorities + percent_uninsured + main_econ + metro_area + (1|state) + (1|region), data=mydata)

anova(model6, model7)
```
percentile_rank_social_vulnerability significant at 0.05 threshold

```{#r}
model8 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_fair_or_poor_health + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

anova(model6, model8)
```
percent_uninsured insignificant

```{#r}
model9 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_fair_or_poor_health + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

model10 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

anova(model8, model9)
anova(model8, model10)
```
percent_minorities significant at 0.05 threshold
percent_fair_or_poor_health insignificant

```{#r}
model11 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + proTrump +  percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

model12 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

model13 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

model14 <- lmer(covid_death_rate_log ~ per_capita_income + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)

anova(model10, model11)
anova(model11, model12)
anova(model12, model13)
anova(model13, model14)
```
percent_below_poverty and proTrump insignificant
per_capita_income and GQ_ESTIMATES_2019 very significant

```{r}
model15 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region), data=mydata)
summary(model15)
```

```{r}
model16 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region) + per_capita_income*metro_area, data=mydata)

model17 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income + percent_minorities + percentile_rank_social_vulnerability + main_econ + metro_area + (1|state) + (1|region) + per_capita_income*metro_area + percent_minorities*per_capita_income, data=mydata)

anova(model16, model15)
anova(model17, model16)
```

```{r}
final.model <- model17
summary(final.model)
#print(final.model, corr=F)
```
covid_death_rate_log ~ GQ_ESTIMATES_2019 + per_capita_income +  
    percent_minorities + percentile_rank_social_vulnerability +  
    main_econ + metro_area + (1 | state) + (1 | region) + per_capita_income *  
    metro_area + percent_minorities * per_capita_income
    
# Plots
```{r}
sjPlot::tab_model(final.model, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```
Effects are pretty much all significant: GQ_ESTIMATES_2019, per_capita_income, percent_minorities, percentile_rank_social_vulnerability (positive), main_econ
[nonspecialized] (baseline: farming), per_capita_income*percent_minorities

```{r}
#plot for fixed effects
#axis labels should be in order from bottom to top
#To see the values of the effect size and p-value, set show.values and show.p= TRUE. 
#P-values will only be shown if the effect size values are too
sjPlot::plot_model(final.model, 
                   #axis.labels=c("per capita income:percent minorities"...),
                   show.values=TRUE, show.p=TRUE,
                   title="Effect of Socio-economic Factors on County COVID Death Rates")
```

# Model Diagnostics
```{r}
plot(final.model, type=c("p", smooth))
qqnorm(residuals(final.model))
ggplot(data.frame(lev=hatvalues(final.model), 
       pearson=residuals(final.model, type="pearson")), 
       aes(x=lev, y=pearson)) +
       geom_point() +
       theme_bw()
```
Residaul plot looks good and no obvious heteroskedasticity. Normality is also good. Seems to have some high leverage points.

```{r}
# scale-location plots
plot(final.model, sqrt(abs(resid(.))) ~ fitted(.), type = c("p", "smooth"))
```

# CV
```{r modelCV, warning=FALSE}
#creating a training set
smp_size <- floor(0.8 * nrow(mydata))

set.seed(1)
train_ind <- sample(seq_len(nrow(mydata)), size = smp_size)

train <- mydata[train_ind, ]
test <- mydata[-train_ind, ]
```

```{r}
#obtaining the mean-squared error for training set
pred.tr <- predict(final.model, train)
MSE.tr <- sum((pred.tr - train$covid_death_rate_log)^2)/nrow(train)
MSE.tr
```

```{r}
#obtaining the mean-squared error for test set
pred.te <- predict(final.model, test)
MSE.te <- sum((pred.te - test$covid_death_rate_log)^2)/nrow(test)
MSE.te
```

```{r}
#obtaining the mean-squared error for dataset
pred <- predict(final.model, mydata)
MSE <- sum((pred - mydata$covid_death_rate_log)^2)/nrow(mydata)
MSE
```

```{r 10-fold-CV}
#https://cran.r-project.org/web/packages/groupdata2/vignettes/cross-validation_with_groupdata2.html
set.seed(1) 
train <- fold(train, k = 10)
train <- train %>% arrange(.folds) # order by .folds
```

```{r}
crossvalidate <- function(data, k, model, dependent, random = FALSE){
  # 'data' is the training set with the ".folds" column
  # 'k' is the number of folds we have
  # 'model' is a string describing a linear regression model formula
  # 'dependent' is a string with the name of the score column we want to predict
  # 'random' is a logical; do we have random effects in the model?
  
  # Initialize empty list for recording performances
  performances <- c()
  
  # One iteration per fold
  for (fold in 1:k){
    # Create training set for this iteration
    # Subset all the datapoints where .folds does not match the current fold
    training_set <- data[data$.folds != fold,]
    
    # Create test set for this iteration
    # Subset all the datapoints where .folds matches the current fold
    testing_set <- data[data$.folds == fold,]
    
    ## Train model
    # If there is a random effect,
    # use lmer() to train model
    # else use lm()
    if (isTRUE(random)){
      # Train linear mixed effects model on training set
      model <- lmer(model, training_set, REML=FALSE)
    } else {
      # Train linear model on training set
      model <- lm(model, training_set)
    }
    ## Test model
    # Predict the dependent variable in the testing_set with the trained model
    predicted <- predict(model, testing_set, allow.new.levels=TRUE)

    # Get the Root Mean Square Error between the predicted and the observed
    RMSE <- rmse(predicted, testing_set[[dependent]])

    # Add the RMSE to the performance list
    performances[fold] <- RMSE
  }
  # Return the mean of the recorded RMSEs
  #return(performances)
  #return(c('RMSE' = mean(performances)))
  return(mean(performances))
}
```

```{r}
#library(hydroGOF) # rmse()
set.seed(1)
RMSE.tr <- crossvalidate(train, k=10, final.model, dependent="covid_death_rate_log", random = TRUE)
```

```{r}
MSE.tr <- RMSE.tr^2
MSE.tr
```
MSE: 0.8341988 

```{r}
# Creating the model for the full training set
cv.model <- lmer(final.model, train, REML = FALSE)

# Predict the dependent variable in the test_set with the trained model
predicted <- predict(cv.model, test, allow.new.levels=TRUE)

# Get the Root Mean Square Error between the predicted and the observed
RMSE.te <- rmse(predicted, test[["covid_death_rate_log"]])
```

```{r}
MSE.te <- RMSE.te^2
MSE.te
```
MSE: 0.8273282

```{r}
# Summarize the results
summary(cv.model)
```

```{r}
sjPlot::tab_model(cv.model, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```
GQ_ESTIMATES_2019 no longer significant

```{r}
# Predict the dependent variable in the full data with the trained model
predicted <- predict(cv.model, mydata, allow.new.levels=TRUE)

# Get the Root Mean Square Error between the predicted and the observed
RMSE <- rmse(predicted, mydata[["covid_death_rate_log"]])
```

```{r}
MSE <- RMSE^2
MSE
```
MSE: 0.8023507