---
title: "Hierarchical Model"
author: "Alice Jiang"
date: "11/17/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
#library(nlme)
library(lme4)
library(ggplot2)
library(dplyr)
library(sjPlot) #for plotting lmer and glmer mods
library(groupdata2)
library(hydroGOF) # rmse()
```

```{r}
mydata <- read_csv("covid.csv") 
mydata <- mydata[is.finite(mydata$covid_death_rate_log),]
mydata$name <- as.factor(mydata$name)
mydata$region <- as.factor(mydata$region)
mydata$state <- as.factor(mydata$state)
mydata$govParty <- as.factor(mydata$govParty)
mydata$metro_area <- as.factor(mydata$metro_area)
mydata$main_econ <- as.factor(mydata$main_econ)
mydata$GQ_ESTIMATES_2019 <- log(mydata$GQ_ESTIMATES_2019) 
mydata <- mydata %>%
  filter(GQ_ESTIMATES_2019 != '-Inf') %>%
  mutate(density = total_population/area_sqmi)
mydata <- mydata[,-c(1:3,29:31)]
```

```{r}
summary(mydata)
#summary(mydata$name)
```
$name$: county
demographic factors: $total\_population$, $percent\_fair\_or\_poor\_health$, $per\_capita\_income$
social factors: $percent\_uninsured$, $percent\_minorities$, $percent\_below\_poverty$, $percentile\_rank\_social\_vulnerability$
geographical factors: $region$ and $state$
economic factors: $main\_econ$ and $metro\_area$
political factors: $pro\_Trump$ and $govParty$

$GQ\_ESTIMATES\_2019$: the total number of county residents living in group quarters in 2019 (e.g. nursing homes, residential treatment centers, student housing, religious convents/monasteries, correctional facilities, and hospitals)

```{#r}
ggplot(mydata, aes(x=covid_death_rate_log)) +
  geom_histogram(fill = "cornflowerblue", color = "white") +
  facet_wrap(~region, ncol = 4) +
  labs(title = "Death Rates by Region")

ggplot(mydata, aes(x=covid_death_rate_log)) +
  geom_histogram(fill = "cornflowerblue", color = "white") +
  facet_wrap(~metro_area, ncol = 3) +
  labs(title = "Death Rates by Metro Area")

ggplot(mydata, aes(x=covid_death_rate_log)) +
  geom_histogram(fill = "cornflowerblue", color = "white") +
  facet_wrap(~govParty, ncol = 2) +
  labs(title = "Death Rates by govParty")
```

# EDA
```{#r}
ggplot(mydata, aes(x = region, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by Region")

ggplot(mydata, aes(x = metro_area, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by Metro Area")

ggplot(mydata, aes(x = govParty, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by govParty")

ggplot(mydata, aes(x = main_econ, y = covid_death_rate_log)) +
    geom_boxplot(notch = TRUE, fill = "cornflowerblue", alpha = .7) + 
    labs(title = "Death Rates by Main Economy")
```
Northeast seems to have higher log(death rates).
metro_area, govParty, main_econ do not seem to have significant impacts on death rates.

```{#r}
ggplot(mydata, aes(x = GQ_ESTIMATES_2019, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = per_capita_income, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = proTrump, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_below_poverty, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_fair_or_poor_health, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_minorities, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percent_uninsured, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm")

ggplot(mydata, aes(x = percentile_rank_social_vulnerability, y = covid_death_rate_log)) +
    geom_point(color= "steelblue") +
    geom_smooth(method = "lm") 
```

*Positive association between group quarters and log(death rates)
*Negative association between proTrump and log(death rates).
Positive association between percent_below_poverty and log(death rates)
*Positive association between percent_fair_or_poor_health and log(death rates)
*Positive association between percent_minorities and log(death rates)
Positive association between percentile_rank_social_vulnerability and log(death rates)
No obvious relationship between log(death rates) and percent_uninsured and per capita income


# Linear Mixed Model 
```{#r}
dotchart(sort(xtabs(~ mydata$name)), cex=0.7) 
dotchart(sort(xtabs(~ mydata$state)), cex=0.7) 
dotchart(sort(xtabs(~ mydata$region)), cex=0.7)
```
Distribution of potential random effects variables. There are many $name$ (county), each only contribute one instances). There's many states, with a pretty skewed distribution -> the state effects can be worrying. Finally, there's only few 4 regions.

```{r modelCV, warning=FALSE}
#creating a training set
smp_size <- floor(0.8 * nrow(mydata))

set.seed(1)
train_ind <- sample(seq_len(nrow(mydata)), size = smp_size)

train <- mydata[train_ind, ]
test <- mydata[-train_ind, ]
```

```{r}
model1 <- lmer(covid_death_rate_log ~ . - region - state + (1|region), data = train)
#summary(model1)
sjPlot::tab_model(model1, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```
proTrump, metro_area(non-metro), main_econ(manufacturing, nonspecialized), lon, lat, per capita income, percent_below_poverty, percent_minorities, percent_smokers, percent_fair_or_poor_health

```{r}
model2 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + birthRate + deathRate + domMigrationRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + (1|region), data = train)
sjPlot::tab_model(model2, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```
birthRate, deathRate, proTrump, lon, lat, per capita income, percent_below_poverty, percent_minorities, percent_uninsured, percent_smokers, metro_area(non-metro), main_econ(manufacturing, nonspecialized)

```{#r}
model3 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + domMigrationRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + density + (1|region), data = train) #density
anova(model3, model2)

model4 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + domMigrationRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + netMigrationRate + (1|region), data = train) #netMigrationRate
anova(model4, model2)
```

```{#r}
model5 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + domMigrationRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region), data = train) #percent_fair_or_poor_health significant
anova(model5, model2)
```

```{#r}
model6 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + domMigrationRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region), data = train) 
#deathRate and birthRate both significant
anova(model5, model6)
```

```{#r}
model7 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region), data = train)
#domMigrationRate not significant
anova(model5, model7)
```
```{#r}
model8 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region), data = train)

anova(model1, model2, model8)
```

```{#r}
model9 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region) + percent_uninsured*per_capita_income, data = train) #significnat interaction
anova(model9, model8)
```
```{#r}
model10 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region) + percent_smokers*per_capita_income, data = train) #significnat interaction
anova(model10, model8)
```

```{#r}
model11 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region) + percent_smokers*per_capita_income + percent_uninsured*per_capita_income , data = train) #significnat interactions
anova(model11, model10)
```

```{#r}
model12 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region) + percent_smokers*per_capita_income + percent_uninsured*per_capita_income + percentile_rank_social_vulnerability*percent_fair_or_poor_health, data = train) #significnat interactions
anova(model12, model11)
```
```{#r}
model13 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 - region + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region) + percent_smokers*per_capita_income + percent_uninsured*per_capita_income + percentile_rank_social_vulnerability*percent_fair_or_poor_health + percent_uninsured*percent_smokers, data = train) #significnat interactions
anova(model13, model12)
```

```{#r}
sjPlot::tab_model(model13, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```

```{r}
model14 <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region) +   percentile_rank_social_vulnerability*percent_fair_or_poor_health, data = train)

sjPlot::tab_model(model14, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```

```{r}
final.model <- lmer(covid_death_rate_log ~ GQ_ESTIMATES_2019 + birthRate + deathRate + proTrump + lon + lat + per_capita_income + percent_below_poverty + percent_minorities + percentile_rank_social_vulnerability + percent_uninsured + percent_smokers + metro_area + main_econ + percent_fair_or_poor_health + (1|region) + percentile_rank_social_vulnerability*percent_fair_or_poor_health, data = train)
anova(final.model, model14, model1, model2)
```

```{r}
summary(final.model)
```
covid_death_rate_log ~ GQ_ESTIMATES_2019 + birthRate + deathRate +  
    proTrump + lon + lat + per_capita_income + percent_below_poverty +  
    percent_minorities + percentile_rank_social_vulnerability +  
    percent_uninsured + percent_smokers + metro_area + main_econ +  
    percent_fair_or_poor_health + (1 | region) +  
    percentile_rank_social_vulnerability * percent_fair_or_poor_health
    
# Plots
```{r}
sjPlot::tab_model(final.model, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```


```{r}
#plot for fixed effects
#axis labels should be in order from bottom to top
#To see the values of the effect size and p-value, set show.values and show.p= TRUE. 
#P-values will only be shown if the effect size values are too
sjPlot::plot_model(final.model, wrap.labels = 5,
                   #axis.labels=c("per capita income:percent minorities"...),
                   show.values=TRUE, show.p=TRUE,
                   title="Effect of Socio-economic Factors on County COVID Death Rates")
```
proTrump and percentile_rank_social_vulnerability 

```{r}
ggplot(mydata, aes(x=proTrump, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="steelblue", alpha=0.5) +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "County Death Rates vs. ProTrump by Region")
```

```{r}
ggplot(mydata, aes(x=proTrump, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="steelblue", alpha=0.5) +
   facet_wrap(~state) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "Changes in Death Rates")
```

```{r}
ggplot(mydata, aes(x=percentile_rank_social_vulnerability, y = covid_death_rate_log)) +
   geom_smooth(se=FALSE, method="lm", color="grey") +
   geom_point(color="steelblue", alpha=0.5) +
   facet_wrap(~region) + 
   theme_minimal(base_size = 9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title = "County Death Rates vs. Social Vulnerability by Region")
```


# Model Diagnostics
```{r}
plot(final.model, type=c("p", smooth))
qqnorm(residuals(final.model))
ggplot(data.frame(lev=hatvalues(final.model), 
       pearson=residuals(final.model, type="pearson")), 
       aes(x=lev, y=pearson)) +
       geom_point() +
       theme_bw()
```
Residaul plot looks good and no obvious heteroskedasticity. Normality is also good. Seems to have some high leverage points.

```{r}
# scale-location plots
plot(final.model, sqrt(abs(resid(.))) ~ fitted(.), type = c("p", "smooth"))
```

# CV
```{r}
#obtaining the mean-squared error for training set
pred.tr <- predict(final.model, train)
MSE.tr <- sum((pred.tr - train$covid_death_rate_log)^2)/nrow(train)
MSE.tr
```
0.8584532

```{r}
#obtaining the mean-squared error for test set
pred.te <- predict(final.model, test)
MSE.te <- sum((pred.te - test$covid_death_rate_log)^2)/nrow(test)
MSE.te
```
0.9024763

```{r}
#obtaining the mean-squared error for dataset
pred <- predict(final.model, mydata)
MSE <- sum((pred - mydata$covid_death_rate_log)^2)/nrow(mydata)
MSE
```
0.8672671

```{r 10-fold-CV}
#https://cran.r-project.org/web/packages/groupdata2/vignettes/cross-validation_with_groupdata2.html
set.seed(1) 
train <- fold(train, k = 10)
train <- train %>% arrange(.folds) # order by .folds
```

```{r}
crossvalidate <- function(data, k, model, dependent, random){
  # 'data' is the training set with the ".folds" column
  # 'k' is the number of folds we have
  # 'model' is a string describing a linear regression model formula
  # 'dependent' is a string with the name of the score column we want to predict
  # 'random' is a logical; do we have random effects in the model?
  
  # Initialize empty list for recording performances
  performances <- c()
  
  # One iteration per fold
  for (fold in 1:k){
    # Create training set for this iteration
    # Subset all the datapoints where .folds does not match the current fold
    training_set <- data[data$.folds != fold,]
    
    # Create test set for this iteration
    # Subset all the datapoints where .folds matches the current fold
    testing_set <- data[data$.folds == fold,]
    
    ## Train model
    # If there is a random effect,
    # use lmer() to train model
    # else use lm()
    if (isTRUE(random)){
      # Train linear mixed effects model on training set
      model <- lmer(model, training_set, REML=FALSE)
    } else {
      # Train linear model on training set
      model <- lm(model, training_set)
    }
    ## Test model
    # Predict the dependent variable in the testing_set with the trained model
    predicted <- predict(model, testing_set, allow.new.levels=TRUE)

    # Get the Root Mean Square Error between the predicted and the observed
    RMSE <- rmse(predicted, testing_set[[dependent]])

    # Add the RMSE to the performance list
    performances[fold] <- RMSE
  }
  # Return the mean of the recorded RMSEs
  #return(performances)
  #return(c('RMSE' = mean(performances)))
  return(mean(performances))
}
```

```{r}
set.seed(1)
RMSE.tr <- crossvalidate(train, k=10, final.model, dependent="covid_death_rate_log", random = TRUE)
```

```{r}
MSE.tr <- RMSE.tr^2
MSE.tr
```
MSE: 0.8780328

```{r}
# Creating the model for the full training set
cv.model <- lmer(final.model, train, REML = FALSE)

# Predict the dependent variable in the test_set with the trained model
predicted <- predict(cv.model, test, allow.new.levels=TRUE)

# Get the Root Mean Square Error between the predicted and the observed
RMSE.te <- rmse(predicted, test[["covid_death_rate_log"]])
```

```{r}
MSE.te <- RMSE.te^2
MSE.te
```
MSE: 0.9023038

```{r}
# Summarize the results
summary(cv.model)
```
covid_death_rate_log ~ GQ_ESTIMATES_2019 + birthRate + deathRate +  
    proTrump + lon + lat + per_capita_income + percent_below_poverty +  
    percent_minorities + percentile_rank_social_vulnerability +  
    percent_uninsured + percent_smokers + metro_area + main_econ +  
    percent_fair_or_poor_health + (1 | region)  +  
    percentile_rank_social_vulnerability * percent_fair_or_poor_health
    
```{r}
sjPlot::tab_model(cv.model, show.re.var= TRUE, 
                  #pred.labels =c("(Intercept)", "per capita income:percent minorities"...),
                  dv.labels= "Effect of Socio-economic Factors on County COVID Death Rates")
```


```{r}
# Predict the dependent variable in the full data with the trained model
predicted <- predict(cv.model, mydata, allow.new.levels=TRUE)

# Get the Root Mean Square Error between the predicted and the observed
RMSE <- rmse(predicted, mydata[["covid_death_rate_log"]])
```

```{r}
MSE <- RMSE^2
MSE
```
MSE: 0.8672894

```{r}
# Summary
summary <- data.frame(
  #var = c("MSE"),
  method = c("Polynomial Model", "Hierarchical Model", "GAM"),
  MSE = c(0.8919, 0.9023, 0.8900)
)
summary
```